<center>

# Affordability of Residential Real Estate in Warsaw

---

**R-Bootcamp Project – W.MSCIDS_RB01.H2501**

Alena Ploshchansky · Rita Ahlborn

*Master of Science in Real Estate*

*Minor in Data Science*

Lucerne University of Applied Sciences and Arts

January 28, 2026

---

</center>


## **1. Introduction and Hypothesis**


### 1.1 Background and purpose of the project


---


## **2. Data Description**
Sources and types of data will be explained here. how many collumns/rows, what kind of variables, etc.

```{r libraries, include = FALSE}
# Load required libraries
# Here maybe we should separate libtraries based on usage
library(dplyr)
library(ggplot2)
library(tidyr)
library(readxl)
library(janitor)
library(stringi)
library(stringr)
library(ggmap)
library(tidygeocoder)
library(writexl)
library(readr)

# Libraries for R markdown
library(knitr)
library(rmarkdown)

# To be updated ...

```


```{r data, include = FALSE}
#### Data set ####
## load data and inspection

## Prices data set
prices <- read_excel("Prices.xlsx")
head(prices)
dim(prices)

## Density data set
density <- read_excel("Density.xlsx", sheet="TABLE")
head(density)
dim(density)

## Median wage data set
median_wage <- read_excel("MedianWage.xlsx", sheet="TABLE")
head(median_wage)
dim(median_wage)

## Data check
glimpse(prices)
glimpse(density)
glimpse(median_wage)

```

---


## **3. Data Preparation**
In this chapter we describe the datasets we have as well as prepare it for further
analysis. We clean it, transform it and merge it into one final dataset.




### 3.1 Data Cleaning and transformation
This chapter is dedicated to cleaning and transforming each dataset separately.
After all the manipulations we will have three clean datasets ready for merging.
We will merge them in this chapter as well.
We have 3 datasets: prices, density, median_wage.

#### *3.1.1 Price Data*
Here we check for missing values and clean the data. We remove columns that are not needed.

```{r prices_values, include = FALSE}
## Check for missing values
sum(is.na(prices))
## Remove and change polish letters, convert to lower case
clean_prices <- prices %>%
  clean_names() %>%
  mutate(across(where(is.character),
            ~ tolower(stri_trans_general(., "Latin-ASCII"))))

## Check cleaned data
glimpse(clean_prices)

## Remove unnecessary columns
clean_prices <- clean_prices %>%
  select(-c("zrodlo_informacji", "cena_wartosc", "waluta", "numer_budynku"))

```
Here we create new variables: price per sqm, log price per sqm, etc.
First is geo data
```{r prices_geo, include = FALSE}
## Coordinates data type conversion and save
if (!file.exists("streets_geo.rds")) {

  streets_geo <- clean_prices %>%
    distinct(ulica) %>%
    mutate(address = paste0(ulica, ", Warsaw, Poland")) %>%
    geocode(
      address = address,
      method = "osm",
      lat = latitude,
      long = longitude
    )

  saveRDS(streets_geo, "streets_geo.rds")
}

# Load saved geo data
streets_geo <- readRDS("streets_geo.rds")
```

#### *3.1.2 Wage Data*


#### *3.1.3 Density Data*


### 3.2 Data merging



---


## **4. Data Analysis**


### 4.1 Descriptive Statistics

#### *4.1.1 Summary Statistics*

#### *4.1.2 Distribution and Visualization*

### 4.2 Hypotheses

#### *4.2.1 Hypothesis 1*

Neighborhoods with higher density show lower prices per sqm than neighborhoods with lower density

1. Simple linear regression: log_price_sqm ~ log_density
2. Map ??


#### *4.2.2 Hypothesis 2*

Affordability of residential real estate on the left bank of the Vistula River in Warsaw is lower in comparison to right bank neighborhoods.

1. Affordability calculation
2. T-test
3. Visuals: boxplot of affordability by bank


#### *4.2.3 Hypothesis 3*

Newer buildings are less affordable than older and very old buildings.

1. T-test
2. bla bla

---


## **5. Discussion and Limitations**


---


## **6. Conclusion**
